Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-11-20 papers/histogram/sad-referee-response-2.md

Dear editor,

We appreciate the referees' careful reading of our paper, and feel that their
feedback has led to a significant improvement. We have added new data and a
section for the 31 atom Lennard-Jones cluster. We show specific heat curves for
this test system and show the convergence SAD vs the other method quantitatively
(a quantitative comparison of this kind has not been done in the literature to
our knowledge). We also added a note to the Square-well fluid section
highlighting that this system has not been explored using broad histogram
methods. We believe that we have addressed all of the referees' concerns and per
the first referee's own words 'If that or similar data would be in the paper
there would be no doubt whatsoever that SAD is a "substantial improvement of
existing computational methods".'

@@

...

Response to first referee
# -----------------------------------------------------------------------------#

In the previous review, the first referee acknowledged that  'The paper is very
well written. It is also true that the authors addressed all of my previous
concerns in a satisfactory manner.'

The first referee asked that we simulate (either 'faster and/or reliably')
low-temperature solid-solid transitions in specific Lennard-Jones clusters. The
referee gave LJ135 as an example, but stated that if specific heat capacity
curves were in the paper for LJ135 or 'similar' (we did LJ31 for reasons we
state in the paper) then 'there would be no doubt whatsoever that SAD is a
"substantial improvement of existing computational methods".'
We believe that we have completely satisfied the first referee and thank the referee
for pointing our attention toward simulating Lennard-Jones clusters.

Response to second referee
# -----------------------------------------------------------------------------#
All of the second referee's comments were addressed previously.

### 2019-11-22 papers/histogram/sad.tex
edits in our meeting

\begin{abstract}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  the 31-atom Lennard-Jones cluster and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interest, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus, SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

Flat histogram methods calculate the density of states $D(E)$ for a discrete set
of energies~\cite{wang2001determining, dayal2004performance, troyer2003flat,
trebst2004optimizing}. For this reason, energy binning becomes an important
consideration for systems with a continuum of possible energies.  Energy bins
are typically of uniform size for the entire energy
continuum~\cite{fasnacht2004adaptive}. Some methods such as
AdaWL~\cite{koh2013dynamically} employ a tunable mechanism for controlling the
binning for low entropic states in order to ensure the exploration of all
energies.  We test the methods on both a discrete and continuum energy system.

@@

A significant advantage of SAD over SAMC---which the $1/t$-WL and WL
methods share \emph{after} they have discovered all the energies---is that
the schedule for $\gamma_t$ automatically responds to the choice of bin
size. Although we have not yet thoroughly tested this over a wide range of bin sizes,
SAD should perform similarly over a reasonable range of bin sizes
because $\gamma_t \propto \nicefrac{N_S}{t}$.  As the number of energy
states $N_S$ found increases (fine binning), the time spent $t$ in
each bin will decrease with the effect that the convergence should be
roughly independent of the bin size chosen.  SAMC could be used with a
prefactor $\gamma_0$ to aid in a similar
way~\cite{werlich2015stochastic} but this adds yet another parameter
for the user to choose.

@@

\subsection{Square-well fluid}
As our first test case, we consider the square-well fluid i.e. a
system of particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.
In addition to providing an ideal test-bed for new algorithm development,
to the best of the knowledge of the authors, there are \emph{not} any direct convergence comparison tests in the literature for the variety of flat histogram methods presented here applied to the square-well fluid.
The
potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

@@

In order to explore both the \jpsays{remove? bin-size behavior} and low-temperature
convergence of our algorithm, we chose to study the 31-atom cluster
(LJ31), which is the smallest Lennard-Jones cluster that exhibits
broken ergoidcity and a solid-solid phase transition at a low
temperature~\cite{doye1998thermodynamics,
  martiniani2014superposition}.  Clusters of Lennard-Jones atoms have
been frequently used for testing Monte Carlo
algorithms~\cite{wales1997global, neirotti2000phase,
  frantsuzov2005size, mandelshtam2006multiple}.  The LJ31 cluster
is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\cite{calvo2000entropic,
  calvo2000phase, poulain2006performances,
  martiniani2014superposition}.  In particular, the LJ31 cluster
features a solid-solid phase transition at a temperature of around $T
\approx 0.027\epsilon/k_B$ that is challenging to
converge~\cite{mandelshtam2006multiple}.  Poulain \emph{et al.}
tested a number of variants of the WL algorithm on LJ31 and failed to
converge the spike in heat capacity corresponding to the solid-solid
phase transition~\cite{poulain2006performances}.

@@

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster
over the temperature range from $k_BT=0.01\epsilon$ to
$k_BT=0.40\epsilon$.  There are two peaks, a large peak around $k_BT
\approx 0.3\epsilon$ corresponding to melting temperature, and a
small peak around $T \approx 0.027\epsilon/k_B$ for the solid-solid
transition from a Mackay-to-anti-Mackay (M $\rightarrow$ aM)
transition~\cite{doye1997thermally, hendy2001molecular,
  mandelshtam2006multiple}.  We constrain our atoms within a spherical
box of radius 2.5$\sigma$, as is common in the
literature~\cite{poulain2006performances, mandelshtam2006multiple}.
We note that using a larger box (with radius $5\sigma$) has a large
effect on the melting peak, and a smaller but still significant effect
on the solid-solid transition.  Since our focus is on the convergence
rate, we focused on the smaller box size.  We further restricted all
our simulations to negative energies.

@@

In order to test the low-temperature convergence, we focused on the
heat capacity in the temperature range $0.01\epsilon \le k_BT \le
0.05\epsilon$, which contains the solid-solid phase transition.
%
We compute the heat capacity from the entropy using a standard
canonical analysis, which gives
\begin{align}
  C_V &= 31\times\frac32k_B + \frac1{k_BT^2}\left< (E- \langle
  E\rangle)^2\right>
  \label{eq:cv}
\end{align}
where the first term represents the kinetic energy contribution, and
the second term is the potential energy contribution, with canonical
averages using our microcanonical entropy to compute multiplicities.
%
For our reference heat capacity, we ran a $1/t$-WL simulation with the
energy constrained to a range $-133.53\epsilon\le E\le -110\epsilon$.
We chose this energy range through experimentation, as we found the
convergence rate to be very sensitive to the minimum energy chosen.

%% \jpcom{this is a basically a repeat of above...
%% The heat capacity changes significantly even at these low temperatures
%% when the size of the cavity is increased.  We use a 2.5 cavity because
%% ``everyone else'' does~\cite{poulain2006performances}.  For purposes
%% of testing the convergence of the algorithm, we don't need to solve
%% for the true heat capacity.  Furthermore, the heat capacity at fixed
%% volume is not the most meaningful quantity to predict for a cluster.
%% }
We ran simulations with bin widths of $\Delta E=0.1\epsilon$,
$0.01\epsilon$, and \jpsays{should we mention if we dont show data? $0.001\epsilon$} using each algorithm we studied.
When examining the converged results, we saw that each bin size
measureably different heat capacity at low temperatures. The largest
bin size of $0.1\epsilon$ introducing a maximum error in heat capacity
of more than 20$k_B$ while the intermediate bin size of $0.01\epsilon$
introduced an error of less than $k_B$ when compared with the
converged result at our smallest bin size.  Therefore we chose to
compare our simulations primarily at the intermediate bin size.  We
used a bin size of $0.01\epsilon$ for our reference in all plots in
this paper (including the reference), except where otherwise
specified.

Figure~\ref{fig:lj-cv-error} (inset) shows the maximum error in heat capacity
in the temperature range $0.01\epsilon\le k_BT\le 0.05\epsilon$ as a
function of the number of Monte Carlo moves performed.
%
We explored the convergence behavior of each algorithm at two energy
bin widths ($\Delta E=0.01\epsilon$ and $\Delta E=0.1\epsilon$).  We
were unable to converge the WL or $1/t$-WL algorithms with the larger
bin size.
%
The rate of convergence for the heat capacity seems to be roughly
independent of the bin width.  In contrast (but not shown) the entropy
itself shows much larger statistical fluctuations with smaller bin
sizes.  The canonical averaging process in Eq.~\ref{eq:cv} averages
out those statistical fluctuations, which reduces the penalty of using
a smaller bin size.

@@

We found the convergence of the WL and $1/t$-WL to be incredibly
sensitive to the minimum energy chosen, complicating the process of
obtaining a workable result.  The known ground state energy of this
cluster is $-133.586422\epsilon$~\cite{northby1987structure,
  wales1997global}.  With a minimum energy of $-133.58\epsilon$, none
of our simulations converged in 10$^{12}$ moves.  With minimum
energies of $-133.54\epsilon$, $-133.53\epsilon$, and
$-133.52\epsilon$, the simulations do seem to converge, although the
lowest energy case was significantly slower and never reached as small
an error.  This result highlights the advantage of being able to
specify a temperature range rather than an energy range when computing
properties as a function of temperature.  It is not only more
convenient, but also far more efficient due the the fact that only a
single simulation need be performed.

@@

Finally, we want to mention our results for SAMC, which are not shown
in our plots in order to reduce clutter.  With the right value for
$t_0$ and with the right energy range chosen, SAMC converges as
rapidly as either $1/t$-WL or SAD.  However, with a wrong $t_0$
parameter, the convergence rate is very poor.  In addition, the
optimal $t_0$ parameter depends on the bin width as well as the energy
range chosen, further complicating matters.

@@

\section{Conclusions}
We have introduced a new algorithm, a variant of the SAMC method, which
effectively samples the
energy space corresponding to a desired range of temperatures for a
variety of system sizes.
%
We find that both SAD and $1/t$-WL demonstrate excellent and robust
convergence.
They both converge more rapidly than SAMC, and unlike WL consistently
converge to the correct density of states. SAD requires
the user to specify a temperature range of interest rather than an
energy range of interest as $1/t$-WL does.  For use cases in which
a range of desired temperatures is known, this will make the SAD method
considerably more convenient.

@@

We found that for examining low-energy structures of a Lennard-Jones
cluster the WL and $1/t$-WL algorithms---versions of which had
previously found to be unsuccessful on this problem~\cite{poulain2006performances}---have
convergence properties that are highly dependent on the choice of
energy range to be examined.  This is particularly problematic when
examining the low-temperature heat capacity, where it is difficult to
determine the lowest energy that will have a significant impact at the
temperatures of interest. This is an example of how SAD is more powerful when the range
of energies is not known in advance.

>>>>

\begin{abstract}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  the 31-atom Lennard-Jones cluster and
  compare the convergence behavior of several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interest, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.
  %
  The convergence of the $1/t$-WL method is very sensitive to the energy
  range chosen for the low-temperature heat capacity of the
  Lennard-Jones cluster.
  %
  Thus, SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

Flat histogram methods calculate the density of states $D(E)$ for a discrete set
of energies~\cite{wang2001determining, dayal2004performance, troyer2003flat,
trebst2004optimizing}. For this reason, energy binning becomes an important
consideration for systems with a continuum of possible energies.  Energy bins
are typically of uniform size for the entire energy
continuum~\cite{fasnacht2004adaptive}. Some methods such as
AdaWL~\cite{koh2013dynamically} employ a tunable mechanism for controlling the
binning for low entropic states in order to ensure the exploration of all
energies.  The method introduced in this paper is designed to scale appropriately
as bin size is changed.

@@

A significant advantage of SAD over SAMC---which the $1/t$-WL and WL
methods share \emph{after} they have discovered all the energies---is that
the schedule for $\gamma_t$ automatically responds to the choice of bin
size.
SAD should perform similarly over a reasonable range of bin sizes
because $\gamma_t \propto \nicefrac{N_S}{t}$.  As the number of energy
states $N_S$ found increases (fine binning), the time spent $t$ in
each bin will decrease with the effect that the convergence should be
roughly independent of the bin size chosen.  SAMC could be used with a
prefactor $\gamma_0$ to aid in a similar
way~\cite{werlich2015stochastic} but this adds yet another parameter
for the user to choose.

@@

\subsection{Square-well fluid}
As our first test case, we consider the square-well fluid i.e. a
system of particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.
To date, there have not been any published direct convergence comparison tests
for flat histogram methods applied to the square-well fluid.
The potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

@@

In order to further explore both the
convergence of our algorithm, we chose to study the 31-atom cluster
(LJ31), which is the smallest Lennard-Jones cluster that exhibits
broken ergoidcity and a solid-solid phase transition at a low
temperature~\cite{doye1998thermodynamics,
  martiniani2014superposition}.  Clusters of Lennard-Jones atoms have
been frequently used for testing Monte Carlo
algorithms~\cite{wales1997global, neirotti2000phase,
  frantsuzov2005size, mandelshtam2006multiple}.  The LJ31 cluster
is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\cite{calvo2000entropic,
  calvo2000phase, poulain2006performances,
  martiniani2014superposition}.  In particular, the LJ31 cluster
features a low-temperature solid-solid phase transition that is challenging to
converge~\cite{mandelshtam2006multiple}.  Poulain \emph{et al.}
tested a number of variants on the WL algorithm on LJ31 and failed to
converge the spike in heat capacity corresponding to the solid-solid
phase transition~\cite{poulain2006performances}.

@@

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster
over the temperature range from $k_BT=0.01\epsilon$ to
$k_BT=0.40\epsilon$.  There are two peaks, a large peak around $k_BT
\approx 0.3\epsilon$ corresponding to melting temperature, and a
small peak around $k_BT \approx 0.027\epsilon$ for the solid-solid
transition from a Mackay-to-anti-Mackay (M $\rightarrow$ aM)
transition~\cite{doye1997thermally, hendy2001molecular,
  mandelshtam2006multiple}.  We constrain our atoms within a spherical
box of radius 2.5$\sigma$, as is common in the
literature~\cite{poulain2006performances, mandelshtam2006multiple}.
We note that using a larger box (with radius $5\sigma$) has a large
effect on the melting peak, and a smaller but still significant effect
on the solid-solid transition.  Since our focus is on the convergence
rate, we will restrict ourselves to the smaller box size.  In addition, we
further restricted all our simulations to negative total potential energies.

@@

In order to test the low-temperature convergence, we focused on the
heat capacity in the temperature range $0.01\epsilon \le k_BT \le
0.05\epsilon$ (see inset in Fig.~\ref{fig:lj-cv}), which contains
the solid-solid phase transition.
%
We compute the heat capacity from the entropy using a standard
canonical analysis, which gives
\begin{align}
  C_V &= 31\times\frac32k_B + \frac1{k_BT^2}\left< (E- \langle
  E\rangle)^2\right>
  \label{eq:cv}
\end{align}
where the first term represents the kinetic energy contribution, and
the second term is the potential energy contribution, with canonical
averages using the microcanonical entropy to compute multiplicities.
%
For our reference heat capacity, we ran a $1/t$-WL simulation with the
energy constrained to a range $-133.53\epsilon\le E\le -110\epsilon$.
We chose this energy range through experimentation, as we found the
convergence rate to be very sensitive to the minimum energy chosen.

%% \jpcom{this is a basically a repeat of above...
%% The heat capacity changes significantly even at these low temperatures
%% when the size of the cavity is increased.  We use a 2.5 cavity because
%% ``everyone else'' does~\cite{poulain2006performances}.  For purposes
%% of testing the convergence of the algorithm, we don't need to solve
%% for the true heat capacity.  Furthermore, the heat capacity at fixed
%% volume is not the most meaningful quantity to predict for a cluster.
%% }
We ran simulations with several bin widths using each algorithm we studied
and show results for $\Delta E=0.1\epsilon$ and
$0.01\epsilon$.
When examining the converged results, we saw that each bin size yielded a
measureably different low temperature heat capacity. The largest
bin size of $0.1\epsilon$ introduced a maximum error in heat capacity
of more than $20k_B$ compared with the smaller bin size.  We will focus on
the smaller bin size, and
used a bin size of $0.01\epsilon$ for our converged reference heat capacity.

Figure~\ref{fig:lj-cv-error} shows the maximum error in heat capacity
over the temperature range $0.01\epsilon\le k_BT\le 0.05\epsilon$ as a
function of the number of Monte Carlo moves performed.
%
We explored the convergence behavior of each algorithm at two energy
bin widths ($\Delta E=0.01\epsilon$ and $\Delta E=0.1\epsilon$).  We
were unable to converge the WL or $1/t$-WL algorithms with the larger
bin size.
%
The rate of convergence of SAD for the heat capacity seems to be roughly
independent of the bin width.  In contrast (but not shown) the entropy
itself shows much larger statistical fluctuations with smaller bin
sizes.  The canonical averaging process in Eq.~\ref{eq:cv} averages
out those statistical fluctuations, which reduces the penalty of using
a smaller bin size.

@@

The convergence behavior of the WL and $1/t$-WL is incredibly
sensitive to the minimum energy chosen, which complicates the process of
obtaining a workable result.  The known ground state energy of this
cluster is $-133.586422\epsilon$~\cite{northby1987structure,
  wales1997global}.  With a minimum energy of $-133.58\epsilon$, none
of our WL or $1/t$-WL simulations converged in 10$^{12}$ moves.  With minimum
energies of $-133.54\epsilon$, $-133.53\epsilon$, and
$-133.52\epsilon$, the simulations do seem to converge, although the
lowest energy case was significantly slower.
This result highlights the advantage of being able to
specify a temperature range rather than an energy range when computing
properties as a function of temperature.  It is not only more
convenient, but also far more efficient due the the fact that only a
single simulation need be performed.

@@

Finally, we want to mention our results for SAMC, which are not shown
in our plots in order to reduce clutter.  With a well-chosen power of ten value for
$t_0$, SAMC can converge to the correct result, but less rapidly than
either $1/t$-WL or SAD.  However, with a poor $t_0$
parameter, the method cannot converge in practice.  In addition, the
optimal $t_0$ parameter depends on the bin width as well as the energy
range chosen, further complicating matters.

@@

\section{Conclusions}
We have introduced a new algorithm, a variant of the SAMC method, which
effectively samples the
energy space corresponding to a desired range of temperatures for a
few systems.
%
We find that both SAD and $1/t$-WL converge more rapidly than SAMC, and
unlike WL consistently converge to the correct density of states. SAD requires
the user to specify a temperature range of interest rather than an
energy range of interest as $1/t$-WL does.  For use cases in which
a range of desired temperatures is known, this will make the SAD method
considerably more convenient.

@@

Finally, we find that for examining a Lennard-Jones
cluster the WL and $1/t$-WL algorithms---versions of which had
previously found to be unsuccessful on this problem~\cite{poulain2006performances}---have
convergence properties that are highly dependent on the choice of
energy range to be examined.  This is particularly problematic when
examining the low-temperature heat capacity, where it is difficult to
determine the lowest energy that will have a significant impact at the
temperatures of interest. This is an example of how SAD is more powerful when the range
of energies is not known in advance.
